---
title: 'kaggle tutorial: titanic survival'
author: "Adam Kowalczewski"
date: "27 July 2016"
output: html_document
---

```{r}
library(readr)
library(rpart)
library(rattle)
library(rpart.plot)
library(RColorBrewer)
setwd("~/R101/titanic/")

train <- read_csv("train.csv")
test <- read_csv("test.csv")

# prediction 1: everyone dies
test$Survived <- rep(0, 418)
submit <- data.frame(PassengerId = test$PassengerId, Survived = test$Survived)
write.csv(submit, file = "theyallperish.csv", row.names = F)
```

# prediction 2: women survive, men die

We can see that most women survived, while very few men did.  We can update our prediction to be that all women survive and see how our score changes.

```{r}
prop.table(table(as.factor(train$Sex), train$Survived),1)

test$Survived[test$Sex == 'female'] <- 1
submit <- data.frame(PassengerId = test$PassengerId, Survived = test$Survived)
#write.csv(submit, file = "womenfirst.csv", row.names = F)
```

# prediction 3: children survive, adults less so

We know the famous "Women and children first" approach to rescue the Titanic took so we can try using the age variable to get some more accuracy.

```{r}
summary(train$Age)

# assume the NAs were the average age (so will not be assigned child below)
train$Child <- 0
train$Child[train$Age < 18] <- 1

# see survivor rate for different subsets
# aggregate takes target variable left of ~ and subsets by vars on the right
aggregate(Survived ~ Child + Sex, data = train, FUN=sum)
# find totals for each subset
aggregate(Survived ~ Child + Sex, data = train, FUN=length)

# combine the two into a proportion
aggregate(Survived ~ Child + Sex, data = train, FUN=function(x){sum(x)/length(x)})



```

It looks like the gender dimension still explains most of the difference.  Being a child didn't rescue males all that much more... [or did it]?

# prediction 4: higher socio economic status increased your chance of survival
We'll bin the continuous variable Fare so we can tabulate it.  We'll use less than $10, $10 - $20, $20 - $30, and more than $30.

It turns out that women in 3rd class who paid for a more expensive ticket were less likely to survive, perhaps because expensive cabins were futher from lifeboats?  We can update our prediction based on this.

```{r}
train$Fare2 <- '30+'
train$Fare2[train$Fare >= 20 & train$Fare < 30] <- '20 - 30'
train$Fare2[train$Fare >= 10 & train$Fare < 20] <- '10 - 20'
train$Fare2[train$Fare < 10] <- '<10'

aggregate(Survived ~ Fare2 + Pclass + Sex, data = train, FUN=function(x){sum(x) / length(x)})
```

```{r}
test$Survived <- 0
test$Survived[test$Sex == 'female'] <- 1
test$Survived[test$Sex == 'female' & test$Pclass == 3 & test$Fare >= 20] <- 0

submit <- data.frame(PassengerId = test$PassengerId, Survived = test$Survived)
write.csv(submit, file = "women_class_combo_model.csv", row.names = F)

```

# prediction 5: using a decision tree

```{r}
str(train)
fit <- rpart(Survived ~ Pclass + Sex + Age + SibSp + Parch + Fare + Embarked,
             data = train,
             method = "class")
plot(fit)
text(fit)
```






